\large STRATEGIC SOCIAL LEARNING AND THE POPULATION DYNAMICS OF HUMAN BEHAVIOR: THE GAME OF GO

### 2018-10-13 4_prepare_manuscript/tex/manuscript.tex
moved the text to Rmarkdown

---
title: "STRATEGIC SOCIAL LEARNING AND THE POPULATION DYNAMICS OF HUMAN BEHAVIOR: THE GAME OF GO"
short_title: "STRATEGIC SOCIAL LEARNING"
date: March 28, 2014
author: Bret A. Beheim$^1$, Calvin Thigpen$^2$, Richard McElreath$^3$
email: beheim@gmail.com
address: $^1$Department of Anthropology, University of New Mexico. $^2$Institute for Transportation Studies, University of California, Davis. $^3$ Department of Anthropology, University of California, Davis
abstract: "Human culture is widely believed to undergo evolution, via mechanisms rooted in the nature of human cognition. A number of theories predict the kinds of human learning strategies, as well as the population dynamics that result from their action. There is little work, however, that quantitatively examines the evidence for these strategies and resulting cultural evolution within human populations. One of the obstacles is the lack of individual-level data with which to link transmission events to larger cultural dynamics. Here, we address this problem with a rich quantitative database from the East Asian board game known as Go. We draw from a large archive of Go games spanning the last six decades of professional play, and find evidence that the evolutionary dynamics of particular cultural variants are driven by a mix of individual and social learning processes. Particular players vary dramatically in their sensitivity to population knowledge, which also varies by age and nationality. The dynamic patterns of opening Go moves are consistent with an ancient, ongoing arms race within the game itself."
---

# Introduction

One of the greatest contributions of evolutionary theory is the ability to answer "why" questions about living things, explaining the existence of their intricate and beautiful designs as adaptations via natural selection. This success extends to study of our own species; we now know a great deal about our behavior, psychology and, more indirectly, our ancestry, as a consequence of evolutionary theory \citep{Barkow1992adaptedmind, LalandBrown2002sense}. Many pathological or idiosyncratic human social phenomena, for example, can be understood as a mismatch between the environment our minds and bodies are "expecting" (subsistence foraging and horticulture) and the one many of us live in (industrial food markets and sedentary lifestyles).

@@

As with genetic evolution, the driving processes are organized within a set of general categories we call "forces", but unlike natural selection, genetic drift, meiotic drive, etc., the forces of cultural evolution are largely based on the details of human psychology. What humans find easy or appealing to learn, what cues and associations they find salient, and how readily certain cultural traits encourage their own social transmission. As a result, much research in cultural evolution has concentrated on human social and individual learning \citep{Mesoudi2006}, and has shown that simple learning heuristics, such as "when in doubt, copy what the majority is doing" are plausible explanations for how humans efficiently identify fitness-enhancing behaviors \citep{McElreathetalSuccessbias2008}.

These learning processes, in aggregate, change the population distribution of cultural variants in a recursive fashion, and in turn alter the environment for future social learning. Thus, by applying what Ernst Mayr called "population thinking" to human culture, it is possible to develop mechanistic theories of how socially-transmitted technologies, beliefs and behaviors spread and decline over time \citep{shennan2009pattern}.

@@

In what follows, we present an analysis of such a high-resolution ($n=31,137$) archive of cultural patterns: records from the East Asian game of Go. Over many decades of professional play, we can see particular strategies and counterstrategies be innovated, copied, decline, and vanish altogether. Our results provide strong evidence that, on the average, these trends are the result of strategic social learning decisions by professional Go players as they adopt and abandon variant openings. This learning model, in aggregate, is the basis for an apparent evolutionary arms race that is largely responsible for the oscillations in Go openings among professional players. In this way, we are able to link individual-level strategic learning to aggregate population dynamics.

# Historical Dynamics in the Game of Go

The East Asian board game known in the West as Go is, by players, one of the most popular in the world, and certainly one of the oldest living games. Originating in China between two and three thousand years ago (where it is called \textit{weiqi}), it has since diffused to Korea (there the game is called \textit{baduk}) and Japan (\textit{go} or \textit{igo}). Through the Japanese, it became widely known in the West in the early 20th century, and most American and European Go terminology is derived from Japanese. With many centuries of paid professional leagues and an abundance of game records (see Fig. 1), Go is an ideal "model system" for the quantitative study of human behavioral change \citep{gobet2004moves}.

@@

Beyond this basic outline of the game, there are two key rules regulating play. First, stones or groups of stones of one player that become completely surrounded on the grid by stones of the other are removed from the board (this is the only time stones are moved, once placed). Second, to prevent the possibility of the game entering an infinite loop, the *ko* rule restricts play so that the same board state cannot appear twice in one game.

@@

The first few moves of game of Go, which allow the most possibilites, are actually the most predictable. The opening period of the game is often characterized by highly normative, canalized play, where players often follow one of a set of memorized move sequences called *joseki* (or *jungsuk* in Korea), thought to grant neither player a major advantage. Once a particular *joseki* pattern has started, professional players will draw on an encyclopedic knowledge of explored routes in deciding their next move.

Despite the conservatism of opening play in professional games, novel variations appear regularly and, if widely copied, become new *joseki* in the canon. Go historian Peter Shotwell writes of the evolution of *joseki*: "Some are hundreds (or thousands) of years old; others are born, die, and then are reborn again as the style of play changes; still others are yet to be born and patiently wait their turn to be discovered in some flash of ingenuity or desperation." \citep{shotwell2011go}.

@@

Subsequent to Takemiya's arrival, the success of the Fourfour appears to have prompted an innovation race in White's Move 2 response. Before 1970, White's response to the Fourfour was almost always a Threefour on an adjacent corner. If we imagine the M1 Fourfour is played in the upper right quadrant (SGF coordinate *pd*), nearly 80\% of the time the M2 response in the late 1960s was coordinate *dc* (Fig. 3). By 1977, at the Fourfour's first peak, the *dc* response declined to about 30\% of games. In the interim, M2 variants at *dp*, *cq* and *dd* all became established in rapid sequence. As measured by the proportion of games White won, *dp* and *cq* both consistently outperformed *dc* in the 1970s, blunting the Fourfour's effectiveness and anticipating its decline in use after 1977.

## Strategic Social Learning

@@

# Methods

We model the decision to use the Fourfour variant at M1 via a binomial model with a logistic link function of the probability of use, fitted by Hamiltonian MCMC in the STAN software package \citep{stan2013stan} and the R programming language. For each game $i$, we model whether Move 1 was a Fourfour ($y_i=1$) or not ($y_i=0$) via a multilevel logistic regression framework. We used two classes of predictors, "Personal Knowledge" and "Population Knowledge". Personal variables include a player's recent (last two years) use of the Fourfour for Move 1, and how often they won games with it, relative to how often they won games without it. This model corresponds to the hypothesis that players update their views of the move, based on how familiar it is to them and how many games they have won using it. "Population Knowledge" models use various predictors constructed from the recent population use of the Fourfour, and the population win rate of the Fourfour relative to its alternatives. These models attempt to measure frequency-dependent and success-biased social learning described above.

@@

# Results

@@

With all predictors centered on means, the intercept estimates that an average Go player will use the Fourfour about 53.2\% of the time during the study period (i.e. logit(.532)=0.13). A player's recent use of a particular opening move strongly predicts their future use; for an otherwise-average player who used the Fourfour about 60\% of their recent games, the probability they will use the Fourfour rises to 58.7\% (OR: 9.39). There is a smaller (OR: 2.77), but unambiguous, positive performance effect from past success with the Fourfour, adding about two percentage points to the probability of use for a player who has won 10\% more games using the Fourfour than his overall win rate. These two personal effects also interact, so that past success enhances the predictive power of past use (or vis versa).

Most important to the question of social learning and cultural evolution, though, is whether there is predictive value in the recent behavior of *other* players, controlling for individual behavior. Indeed, when the Fourfour appears in 60\% of *all* recent games, the expected probability of use by an average player rises from 53.2\% to 59.0\% (OR: 10.475). Two important interaction effects regulate the frequency-dependence of a player's behavior; if the Fourfour appears in 60\% of recent professional games, *and* its population performance is ten percentage points higher than that of its rival, the Threefour, the expected probability a player will use it rises from 59.0\% to 61.4\% (Fig. 4).

Because the coefficient for personal use (which we call $\beta$) and that for population use ($\gamma$) have the same units, comparing their magnitudes tells us the relative predictive value of personal and population knowledge. For an average player, social knowledge is 1.05 times more important than personal experience in predicting the outcome, but this belies the enormous variation from player to player. Partially pooling games of each Black player, we see a varying effects standard deviation of 4.00 for the distribution of player-specific $\beta_j$. For population knowledge, the standard deviation across player intercepts is even larger, 9.77. Plotting the joint posterior densities of $\beta_j$ and $\gamma_j$ for each player $j$ (Fig. 5, left) reveals that some professionals are substantially better predicted by social knowledge than by their own experiences, while other players are almost completely insensitive to population trends (foremost among them, Takemiya Masaki).

Including player age as a varying effect regulating magnitude of $\beta$ and $\gamma$ also shows learning heterogeneity over the life course (Fig. 5, right). The games of the youngest professionals, those between the ages of 14 and 18, are least predicted by personal experience and most by population trends. After age 20, the balance steadily shifts towards personal experience, peaking at age 27 well above the average $\beta$ of 2.24. Players older than 35 appear to return to average levels of social learning; with a few deviations, no trend is apparent after this age.

# Discussion

@@

According to player-specific posterior estimates in Fig. 5, the players who most strongly track population trends are South Korean professionals Lee Sedol and Cho Namchul, while Chinese player Peng Quan and Japanese player Takemiya Masaki lie at the other extreme. This segregation of nationalities may not be a coincidence; as Table 2 shows, the average of the player-specific estimates for individual and population knowledge differs across country, with South Korean players tending toward greater-than-average sensitivity to social information (positive mean $\gamma_j$). The average Japanese player, in contrast, is better predicted by their individual knowledge (positive mean $\beta_j$), with the 75 Chinese players nearest to the population intercepts $\beta$ and $\gamma$, on the average. It is important to emphasize that these distributions show substantial overlap, but this result raises the tantalizing possibility that normative or institutional differences across East Asia may encourage different forms of social learning within each national league.

@@

Another problem is the nature of innovation in Fourfour play. Even if Takemiya's innovative opening style (widely known as "Cosmic Go") was the catalyst for the Fourfour's rise in the 1970s, as the evidence above seems to indicate, we do not know if other adopters were using it in the same way. It could be, for example, that Takemiya's early successes in 1968-1970 encouraged other professionals to revisit the Fourfour and develop their own innovative applications, rather than emulate his.

@@

\bibliographystyle{apalike}
\bibliography{citations}

@@

# Appendix

## The GoGoD Database

@@

Since we are using player-specific predictors and grouping games by player in our varying effects, we excluded games whose Black players did not have at least 50 games on record between 1956 and 2009. After these exclusions, a total of 31,137 games remained within the database, with 207 unique players taking Black. In addition to the data contained within the game record and the metadata, we also collected player nationality and date of birth was drawn from a variety of sources, chiefly player biography pages from the Go wiki "Sensei's Library" (http://senseis.xmp.net/). Although the GoGoD database is itself commercially licensed by Fairbairn and Hall, the regression table used here is available by request to the corresponding author.

## Full Regression Model

Our basic question is whether or not a player will use the Fourfour opening move in any particular game, modelled as a binomial random variable with a logistic probability model. This regression model has three important features: terms relating to personal information, terms relating to population information, and stochastically modelled player-specific intercepts and slopes. Specifically, if the probability of using the Fourfour in game $i$ of Black player $j$ who is aged $k$ is $p_{ijk}$, then the log-odds that the Fourfour opening move is used is modelled by

@@

Predictor $x_i$ is the proportion of Black player $j$'s games that use the M1 Fourfour recently (within the last two years), $y_i$ is the frequency of the M1 Fourfour among recent games within the general population, $q_i$ is that player's recent win rate specifically among their Fourfour games, and $r_i$ is the win rate among all recent Fourfour games in the population. Of particular interest are parameters $\beta_{ijk}$ and $\gamma_{ijk}$, which represent the importance of personal and population use frequencies in the model. These two parameters are themselves modelled by
$$
  \beta_{ijk} = \beta + \beta_j + \beta_k + \beta_{S} s_{i} + \beta_{Q} q_{i}
$$
$$
  \gamma_{ijk} = \gamma + \gamma_j + \gamma_k + \gamma_{S} s_{i} + \gamma_{R} r_{i}
$$

In addition to interactions for $q_i$ and $r_i$, which allow the Fourfour's recent win rate for the player and the population to regulate the importance of its frequency, predictor $s_i$ is the Black player's overall win rate among games played in the last two years. Within the model, intercept $\alpha_j$ and slopes $\beta_j$ and $\gamma_j$ for player $j$ treated as varying effects following a joint normal distribution with mean \textbf{0}, as are those of players of age $k$.

The controls for each game $i$ are made up of six predictors: White's win rate using the M1 Fourfour in past games (where they played as Black), White's loss rate \textit{against} the M1 Fourfour, White's frequency of use of the Fourfour, the number of past games Black has seen the Fourfour used, whether or not Black lost the last time someone used an M1 Fourfour on him, and the handicap of game $i$.

This model was first fit using the lme4 package in the R language and by Hamiltonian Monte Carlo via the STAN software package called by the glmer2stan command (McElreath 2013), running 5,000 iterations with a 1,000 iteration burn-in. Convergence of chains was determined by the R-hat Gelman & Rubin diagnostic \citep{stan2013stan} and by visual inspection of the chains.

>>>>

%Notes to self on compiling without BibTeX
%Compile original file as normal with BibTex, using bibunits
%copy contents of bu1.bbl and paste into bibliography section
%comment out \usepackage{bibunits} in preamble
%comment out begin and end bibunit, and putbib
%compile twice with XeLateX

%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode
\documentclass[reqno,12pt]{amsart}
\usepackage[foot]{amsaddr}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[paperwidth=8.5in,paperheight=11in,text={6in,9in},left=1.25in,top=1in,headheight=0.25in,headsep=0.4in,footskip=0.4in]{geometry}

%citation format
\usepackage[authoryear]{natbib}
%use authoryear in brackets for conventional format
% use "super" in brackets for numbers to biblio

% \bibpunct[, ]{}{}{,}{s}{}{,}
\bibpunct[, ]{(}{)}{,}{a}{}{,}
%change s to a, and put open and closed parentheses in first two bracketed arguments for author-year citations

\usepackage{subfigure}
\usepackage{lineno}
\usepackage{threeparttable}

\usepackage{comment} %make block comments

\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\E}{E}

%get rid of red link boxes
\usepackage{hyperref}
%\usepackage{endfloat}
\hypersetup{
    colorlinks,
    linkcolor={red!80!black},
    citecolor={blue!90!black},
    urlcolor={blue!90!black}
}

%flexible hyperrefs
%\usepackage{cleveref}

%gives ability to fix figure locations with H
\usepackage{float}

%allows subsetting of bibliography units
%\usepackage{bibunits}

\synctex=1

\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
  \renewenvironment{#1}%
     {\linenomath\csname old#1\endcsname}%
     {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

%\usepackage{lmodern}
%\usepackage{unicode-math}
\usepackage{amsmath}
\usepackage{mathspec}
\usepackage{units}
\usepackage{xltxtra}
\usepackage{xunicode}
\defaultfontfeatures{Mapping=tex-text}
%\setmainfont[Ligatures={Common}]{EB Garamond}
%\setmathrm{EB Garamond}
%\setmathfont(Digits,Latin)[Numbers={Lining,Proportional}]{EB Garamond}

\definecolor{linenocolor}{gray}{0.5}
\renewcommand\thelinenumber{\color{linenocolor}\arabic{linenumber}}

\usepackage{fix-cm}

\setcounter{totalnumber}{10}

\newcommand{\mr}{\mathrm}
\usepackage{hyperref}
\usepackage[normalem]{ulem}
\useunder{\uline}{\ul}{}
\makeatletter
\newenvironment{startquote}[2][2em]
  {\setlength{\@tempdima}{#1}%
   \def\chapquote@author{#2}%
   \parshape 1 \@tempdima \dimexpr\textwidth-2\@tempdima\relax%
   \itshape}
  {\par\normalfont\hfill--\ \chapquote@author\hspace*{\@tempdima}\par\bigskip}
\makeatother
%\renewcommand{\baselinestretch}{1.5}

\title[STRATEGIC SOCIAL LEARNING]
{\large STRATEGIC SOCIAL LEARNING AND THE POPULATION DYNAMICS OF HUMAN BEHAVIOR:
THE GAME OF GO}
\author{Bret A. Beheim\(^1\), Calvin Thigpen\(^2\), Richard McElreath\(^3\)}
\address{\(^1\)Department of Anthropology, University of New Mexico.
\(^2\)Institute for Transportation Studies, University of California,
Davis. \(^3\) Department of Anthropology, University of California,
Davis}
\email{beheim@gmail.com}
\date{March 28, 2014}

\begin{document}
%\begin{bibunit}[naturemag]

%{\abstract \small . }

%\renewcommand*\rmdefault{Minion}

\linenumbers
\modulolinenumbers[5]

\maketitle

%ABSTRACT
%\begin{abstract}\normalsize
\noindent\textsc{Abstract.}
Human culture is widely believed to undergo evolution, via mechanisms
rooted in the nature of human cognition. A number of theories predict
the kinds of human learning strategies, as well as the population
dynamics that result from their action. There is little work, however,
that quantitatively examines the evidence for these strategies and
resulting cultural evolution within human populations. One of the
obstacles is the lack of individual-level data with which to link
transmission events to larger cultural dynamics. Here, we address this
problem with a rich quantitative database from the East Asian board game
known as Go. We draw from a large archive of Go games spanning the last
six decades of professional play, and find evidence that the
evolutionary dynamics of particular cultural variants are driven by a
mix of individual and social learning processes. Particular players vary
dramatically in their sensitivity to population knowledge, which also
varies by age and nationality. The dynamic patterns of opening Go moves
are consistent with an ancient, ongoing arms race within the game
itself.
%\end{abstract}

\vspace{12pt}

\vspace{24pt}
% \newpage

\section{Introduction}\label{introduction}

One of the greatest contributions of evolutionary theory is the ability
to answer ``why'' questions about living things, explaining the
existence of their intricate and beautiful designs as adaptations via
natural selection. This success extends to study of our own species; we
now know a great deal about our behavior, psychology and, more
indirectly, our ancestry, as a consequence of evolutionary theory
\citep{Barkow1992adaptedmind, LalandBrown2002sense}. Many pathological
or idiosyncratic human social phenomena, for example, can be understood
as a mismatch between the environment our minds and bodies are
``expecting'' (subsistence foraging and horticulture) and the one many
of us live in (industrial food markets and sedentary lifestyles).

@@

As with genetic evolution, the driving processes are organized within a
set of general categories we call ``forces'', but unlike natural
selection, genetic drift, meiotic drive, etc., the forces of cultural
evolution are largely based on the details of human psychology. What
humans find easy or appealing to learn, what cues and associations they
find salient, and how readily certain cultural traits encourage their
own social transmission. As a result, much research in cultural
evolution has concentrated on human social and individual learning
\citep{Mesoudi2006}, and has shown that simple learning heuristics, such
as ``when in doubt, copy what the majority is doing'' are plausible
explanations for how humans efficiently identify fitness-enhancing
behaviors \citep{McElreathetalSuccessbias2008}.

These learning processes, in aggregate, change the population
distribution of cultural variants in a recursive fashion, and in turn
alter the environment for future social learning. Thus, by applying what
Ernst Mayr called ``population thinking'' to human culture, it is
possible to develop mechanistic theories of how socially-transmitted
technologies, beliefs and behaviors spread and decline over time
\citep{shennan2009pattern}.

@@

In what follows, we present an analysis of such a high-resolution
(\(n=31,137\)) archive of cultural patterns: records from the East Asian
game of Go. Over many decades of professional play, we can see
particular strategies and counterstrategies be innovated, copied,
decline, and vanish altogether. Our results provide strong evidence
that, on the average, these trends are the result of strategic social
learning decisions by professional Go players as they adopt and abandon
variant openings. This learning model, in aggregate, is the basis for an
apparent evolutionary arms race that is largely responsible for the
oscillations in Go openings among professional players. In this way, we
are able to link individual-level strategic learning to aggregate
population dynamics.

\section{Historical Dynamics in the Game of
Go}\label{historical-dynamics-in-the-game-of-go}

The East Asian board game known in the West as Go is, by players, one of
the most popular in the world, and certainly one of the oldest living
games. Originating in China between two and three thousand years ago
(where it is called \textit{weiqi}), it has since diffused to Korea
(there the game is called \textit{baduk}) and Japan (\textit{go} or
\textit{igo}). Through the Japanese, it became widely known in the West
in the early 20th century, and most American and European Go terminology
is derived from Japanese. With many centuries of paid professional
leagues and an abundance of game records (see Fig. 1), Go is an ideal
``model system'' for the quantitative study of human behavioral change
\citep{gobet2004moves}.

@@

Beyond this basic outline of the game, there are two key rules
regulating play. First, stones or groups of stones of one player that
become completely surrounded on the grid by stones of the other are
removed from the board (this is the only time stones are moved, once
placed). Second, to prevent the possibility of the game entering an
infinite loop, the \emph{ko} rule restricts play so that the same board
state cannot appear twice in one game.

@@

The first few moves of game of Go, which allow the most possibilites,
are actually the most predictable. The opening period of the game is
often characterized by highly normative, canalized play, where players
often follow one of a set of memorized move sequences called
\emph{joseki} (or \emph{jungsuk} in Korea), thought to grant neither
player a major advantage. Once a particular \emph{joseki} pattern has
started, professional players will draw on an encyclopedic knowledge of
explored routes in deciding their next move.

Despite the conservatism of opening play in professional games, novel
variations appear regularly and, if widely copied, become new
\emph{joseki} in the canon. Go historian Peter Shotwell writes of the
evolution of \emph{joseki}: ``Some are hundreds (or thousands) of years
old; others are born, die, and then are reborn again as the style of
play changes; still others are yet to be born and patiently wait their
turn to be discovered in some flash of ingenuity or desperation.''
\citep{shotwell2011go}.

@@

Subsequent to Takemiya's arrival, the success of the Fourfour appears to
have prompted an innovation race in White's Move 2 response. Before
1970, White's response to the Fourfour was almost always a Threefour on
an adjacent corner. If we imagine the M1 Fourfour is played in the upper
right quadrant (SGF coordinate \emph{pd}), nearly 80\% of the time the
M2 response in the late 1960s was coordinate \emph{dc} (Fig. 3). By
1977, at the Fourfour's first peak, the \emph{dc} response declined to
about 30\% of games. In the interim, M2 variants at \emph{dp}, \emph{cq}
and \emph{dd} all became established in rapid sequence. As measured by
the proportion of games White won, \emph{dp} and \emph{cq} both
consistently outperformed \emph{dc} in the 1970s, blunting the
Fourfour's effectiveness and anticipating its decline in use after 1977.

\subsection{Strategic Social Learning}\label{strategic-social-learning}

@@

\section{Methods}\label{methods}

We model the decision to use the Fourfour variant at M1 via a binomial
model with a logistic link function of the probability of use, fitted by
Hamiltonian MCMC in the STAN software package \citep{stan2013stan} and
the R programming language. For each game \(i\), we model whether Move 1
was a Fourfour (\(y_i=1\)) or not (\(y_i=0\)) via a multilevel logistic
regression framework. We used two classes of predictors, ``Personal
Knowledge'' and ``Population Knowledge''. Personal variables include a
player's recent (last two years) use of the Fourfour for Move 1, and how
often they won games with it, relative to how often they won games
without it. This model corresponds to the hypothesis that players update
their views of the move, based on how familiar it is to them and how
many games they have won using it. ``Population Knowledge'' models use
various predictors constructed from the recent population use of the
Fourfour, and the population win rate of the Fourfour relative to its
alternatives. These models attempt to measure frequency-dependent and
success-biased social learning described above.

@@

\section{Results}\label{results}

@@

With all predictors centered on means, the intercept estimates that an
average Go player will use the Fourfour about 53.2\% of the time during
the study period (i.e.~logit(.532)=0.13). A player's recent use of a
particular opening move strongly predicts their future use; for an
otherwise-average player who used the Fourfour about 60\% of their
recent games, the probability they will use the Fourfour rises to 58.7\%
(OR: 9.39). There is a smaller (OR: 2.77), but unambiguous, positive
performance effect from past success with the Fourfour, adding about two
percentage points to the probability of use for a player who has won
10\% more games using the Fourfour than his overall win rate. These two
personal effects also interact, so that past success enhances the
predictive power of past use (or vis versa).

Most important to the question of social learning and cultural
evolution, though, is whether there is predictive value in the recent
behavior of \emph{other} players, controlling for individual behavior.
Indeed, when the Fourfour appears in 60\% of \emph{all} recent games,
the expected probability of use by an average player rises from 53.2\%
to 59.0\% (OR: 10.475). Two important interaction effects regulate the
frequency-dependence of a player's behavior; if the Fourfour appears in
60\% of recent professional games, \emph{and} its population performance
is ten percentage points higher than that of its rival, the Threefour,
the expected probability a player will use it rises from 59.0\% to
61.4\% (Fig. 4).

Because the coefficient for personal use (which we call \(\beta\)) and
that for population use (\(\gamma\)) have the same units, comparing
their magnitudes tells us the relative predictive value of personal and
population knowledge. For an average player, social knowledge is 1.05
times more important than personal experience in predicting the outcome,
but this belies the enormous variation from player to player. Partially
pooling games of each Black player, we see a varying effects standard
deviation of 4.00 for the distribution of player-specific \(\beta_j\).
For population knowledge, the standard deviation across player
intercepts is even larger, 9.77. Plotting the joint posterior densities
of \(\beta_j\) and \(\gamma_j\) for each player \(j\) (Fig. 5, left)
reveals that some professionals are substantially better predicted by
social knowledge than by their own experiences, while other players are
almost completely insensitive to population trends (foremost among them,
Takemiya Masaki).

Including player age as a varying effect regulating magnitude of
\(\beta\) and \(\gamma\) also shows learning heterogeneity over the life
course (Fig. 5, right). The games of the youngest professionals, those
between the ages of 14 and 18, are least predicted by personal
experience and most by population trends. After age 20, the balance
steadily shifts towards personal experience, peaking at age 27 well
above the average \(\beta\) of 2.24. Players older than 35 appear to
return to average levels of social learning; with a few deviations, no
trend is apparent after this age.

\section{Discussion}\label{discussion}

@@

According to player-specific posterior estimates in Fig. 5, the players
who most strongly track population trends are South Korean professionals
Lee Sedol and Cho Namchul, while Chinese player Peng Quan and Japanese
player Takemiya Masaki lie at the other extreme. This segregation of
nationalities may not be a coincidence; as Table 2 shows, the average of
the player-specific estimates for individual and population knowledge
differs across country, with South Korean players tending toward
greater-than-average sensitivity to social information (positive mean
\(\gamma_j\)). The average Japanese player, in contrast, is better
predicted by their individual knowledge (positive mean \(\beta_j\)),
with the 75 Chinese players nearest to the population intercepts
\(\beta\) and \(\gamma\), on the average. It is important to emphasize
that these distributions show substantial overlap, but this result
raises the tantalizing possibility that normative or institutional
differences across East Asia may encourage different forms of social
learning within each national league.

@@

Another problem is the nature of innovation in Fourfour play. Even if
Takemiya's innovative opening style (widely known as ``Cosmic Go'') was
the catalyst for the Fourfour's rise in the 1970s, as the evidence above
seems to indicate, we do not know if other adopters were using it in the
same way. It could be, for example, that Takemiya's early successes in
1968-1970 encouraged other professionals to revisit the Fourfour and
develop their own innovative applications, rather than emulate his.

@@

\bibliographystyle{apalike}\bibliography{citations}

@@

\section{Appendix}\label{appendix}

\subsection{The GoGoD Database}\label{the-gogod-database}

@@

Since we are using player-specific predictors and grouping games by
player in our varying effects, we excluded games whose Black players did
not have at least 50 games on record between 1956 and 2009. After these
exclusions, a total of 31,137 games remained within the database, with
207 unique players taking Black. In addition to the data contained
within the game record and the metadata, we also collected player
nationality and date of birth was drawn from a variety of sources,
chiefly player biography pages from the Go wiki ``Sensei's Library''
(http://senseis.xmp.net/). Although the GoGoD database is itself
commercially licensed by Fairbairn and Hall, the regression table used
here is available by request to the corresponding author.

\subsection{Full Regression Model}\label{full-regression-model}

Our basic question is whether or not a player will use the Fourfour
opening move in any particular game, modelled as a binomial random
variable with a logistic probability model. This regression model has
three important features: terms relating to personal information, terms
relating to population information, and stochastically modelled
player-specific intercepts and slopes. Specifically, if the probability
of using the Fourfour in game \(i\) of Black player \(j\) who is aged
\(k\) is \(p_{ijk}\), then the log-odds that the Fourfour opening move
is used is modelled by

@@

Predictor \(x_i\) is the proportion of Black player \(j\)'s games that
use the M1 Fourfour recently (within the last two years), \(y_i\) is the
frequency of the M1 Fourfour among recent games within the general
population, \(q_i\) is that player's recent win rate specifically among
their Fourfour games, and \(r_i\) is the win rate among all recent
Fourfour games in the population. Of particular interest are parameters
\(\beta_{ijk}\) and \(\gamma_{ijk}\), which represent the importance of
personal and population use frequencies in the model. These two
parameters are themselves modelled by \[
  \beta_{ijk} = \beta + \beta_j + \beta_k + \beta_{S} s_{i} + \beta_{Q} q_{i}
\] \[
  \gamma_{ijk} = \gamma + \gamma_j + \gamma_k + \gamma_{S} s_{i} + \gamma_{R} r_{i}
\]

In addition to interactions for \(q_i\) and \(r_i\), which allow the
Fourfour's recent win rate for the player and the population to regulate
the importance of its frequency, predictor \(s_i\) is the Black player's
overall win rate among games played in the last two years. Within the
model, intercept \(\alpha_j\) and slopes \(\beta_j\) and \(\gamma_j\)
for player \(j\) treated as varying effects following a joint normal
distribution with mean \textbf{0}, as are those of players of age \(k\).

The controls for each game \(i\) are made up of six predictors: White's
win rate using the M1 Fourfour in past games (where they played as
Black), White's loss rate \textit{against} the M1 Fourfour, White's
frequency of use of the Fourfour, the number of past games Black has
seen the Fourfour used, whether or not Black lost the last time someone
used an M1 Fourfour on him, and the handicap of game \(i\).

This model was first fit using the lme4 package in the R language and by
Hamiltonian Monte Carlo via the STAN software package called by the
glmer2stan command (McElreath 2013), running 5,000 iterations with a
1,000 iteration burn-in. Convergence of chains was determined by the
R-hat Gelman \& Rubin diagnostic \citep{stan2013stan} and by visual
inspection of the chains.

\end{document}
